### 2018-04-05 paper/2-background.tex
Drop hanging word, tweak whitespace
Tweak whitespace
Add authors, minor changes
Improve Excel wording in section 2

\clearpage
\section{Background}\label{sec-background}

@@

The following property is essential for build systems, it is their raison d'\^etre:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10      B1: IF(C1=1,A1,A2)      C1: 1
A2: 20
\end{minted}
\vspace{1mm}

\noindent
Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, which is
not known statically\footnote{One might say that the value of \cmd{C1} is statically
known in this particular example, but imagine that it is the result of a long computation
chain -- its value will only become available during the build.}.

\Excel handles this example correctly: if \cmd{C1=1} and the user changes
\cmd{A2}, \Excel will not require \cmd{B1} to be computed in advance. Alas, other forms of
dynamic dependencies can force \Excel to perform unnecessary computation.
Consider the following modification of the above example:

@@

\noindent
The new version uses the \cmd{INDIRECT} function, which allows us to reference a
cell indirectly by a text string that is not necessarily known in advance. The
current implementation of \Excel recomputes indirect references in every
build~\cite{excel_recalc}. This approach clearly violates the minimality
property~\ref{def-minimal}: if \cmd{C1=1} and the user modifies \cmd{A2}, \Excel
will recompute \cmd{B1}, potentially triggering further unnecessary
recalculation even though \cmd{B1} does not transitively depend on \cmd{A2}.

\Excel's build algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel uses the \emph{calculation chain}
produced by the previous build as an approximation to the correct topological
order. During recalculation, \Excel processes cells in this order, but can
\emph{defer recalculation of a cell} by moving it down the chain if a newly
discovered dependency has not yet been rebuilt. We refer to this
algorithm as \emph{reordering}, and will discuss it in more detail
in~\S\ref{sec-implementation-excel}.

@@

\begin{figure}[h]
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
\caption{Dependency graph produced after the previous build.}
\end{subfigure}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
\caption{Since \cmd{docs.txt} was modified, we rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering a new dependency.}
\end{subfigure}
\vspace{-2mm}
\caption{Dynamic dependencies example: create \cmd{README} and add it to the
list of release documents \cmd{docs.txt}.\label{fig-shake}}
\vspace{-2mm}
\end{figure}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
"release.tar" %> \_ -> do
    need ["release.txt"]
    files <- lines <$> readFile "release.txt"
    need files
    system "tar" $ ["-cf", "result.tar"] ++ files
\end{minted}
\vspace{1mm}

\noindent
We first declare the static dependency on \cmd{release.txt}, then read its
content (a list of files) and depend on each listed file, dynamically. Finally, we specify the
command to produce the resulting archive. Crucially, the archive will only be
rebuilt if one of the dependencies (static or dynamic) has changed. For example,
if we create another documentation file \cmd{README} and add it to
\cmd{docs.txt}, \Shake will appropriately rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering the new dependency, as illustrated in
Fig.~\ref{fig-shake}(b).

@@

\begin{figure}[t]
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
\vspace{-1mm}
\caption{Checkout source files}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
\vspace{-1mm}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
\vspace{-1mm}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
\vspace{-2mm}
\caption{A cloud build example: (a)~checkout sources, (b)~download \cmd{main.exe}
from the cloud and skip intermediate files (only their hashes are needed),
(c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which requires building
\cmd{util.o} (since nobody has compiled \cmd{util.c} before) and downloading
\cmd{main.o}. File hashes are shown inside circles.
\label{fig-bazel}}
\vspace{-1mm}
\end{figure}

@@

\begin{table}[h]
\vspace{-2mm}
\smaller
\centering
\begin{tabular}{l||l|l||l|c|c|c}
\hline
$\!$Build system$\!$& Persistent build information   & Algorithm    & Dependencies & Minimal & Cutoff & Cloud$\!$\\\hline
$\!$\Make       $\!$& File modification times        & Topological  & Static       & Yes     & No     & No   $\!$\\
$\!$\Excel      $\!$& Dirty cells, calculation chain & Reordering   & Dynamic      & No      & No     & No   $\!$\\
$\!$\Shake      $\!$& Previous dependency graph      & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
$\!$\Bazel      $\!$& All dependency graphs, cache   & Topological  & Static       & Yes     & Yes    & Yes  $\!$\\\hline
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Summary of build system differences.\label{tab-summary}}
\vspace{-8mm}
\end{table}

>>>>

\section{Background}\label{sec-background}

@@

The following property is essential for build systems; indeed, it is their raison d'\^etre:

@@

@@

\noindent
The formula in \cmd{B1} uses the \cmd{INDIRECT} function, which takes a string and
returns the value of the cell with that name.  The string evaluates
to \cmd{"A1"}, so \cmd{B1} evaluates to \cmd{10}. However, the dependencies of the
formula in \cmd{B1} are determined by the value of \cmd{C1}, so there is no
possibility\footnote{In this particular example one might say that
the value of \cmd{C1} is statically known, but imagine that it is the result of a long computation
chain -- its value will only become available during the build.}
of computing the dependency graph prior to calc.

To support dynamic dependencies,
\Excel's build algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel uses the \emph{calculation chain}
produced by the previous build as an approximation to the correct topological
order. During recalculation, \Excel processes cells in this order, but can
\emph{defer recalculation of a cell} by moving it down the chain if a newly
discovered dependency has not yet been rebuilt. We refer to this
algorithm as \emph{reordering}, and will discuss it in more detail
in~\S\ref{sec-implementation-excel}.

Dynamic dependencies complicate minimality.  In the above example,
\cmd{B1} should only be recomputed if \cmd{A1} or \cmd{C1} change, but
not if (say) \cmd{A2} changes; but these facts are not statically apparent.
In pracice Excel implements a conservative approximation to
minimality: roughly, it recomputes a formula if (a) the formula
text mentions a changed cell, or (b) the formula uses a function like \cmd{INDIRECT}
whose dependencies are not statically visible.

@@

\begin{figure}[h]
\vspace{-2mm}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
\caption{Dependency graph produced after the previous build.}
\end{subfigure}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
\caption{Since \cmd{docs.txt} was modified, we rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering a new dependency.}
\end{subfigure}
\vspace{-2mm}
\caption{Dynamic dependencies example: create \cmd{README} and add it to the
list of release documents \cmd{docs.txt}.\label{fig-shake}}
\vspace{-4mm}
\end{figure}

@@

\begin{minted}[xleftmargin=10pt]{haskell}
"release.tar" %> \_ -> do
    need ["release.txt"]
    files <- lines <$> readFile "release.txt"
    need files
    system "tar" $ ["-cf", "result.tar"] ++ files
\end{minted}

\noindent
We first declare the static dependency on \cmd{release.txt}, then read its
content (a list of files) and depend on each listed file, dynamically. Finally, we specify the
command to produce the resulting archive. Crucially, the archive will only be
rebuilt if one of the dependencies (static or dynamic) has changed. For example,
if we create another documentation file \cmd{README} and add it to
\cmd{docs.txt}, \Shake will appropriately rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering the new dependency, see Fig.~\ref{fig-shake}(b).

@@

\begin{figure}[t]
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
\caption{Checkout source files}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
\caption{A cloud build example: (a)~checkout sources, (b)~download \cmd{main.exe}
from the cloud and skip intermediate files (only their hashes are needed),
(c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which requires building
\cmd{util.o} (since nobody has compiled \cmd{util.c} before) and downloading
\cmd{main.o}. File hashes are shown inside circles.
\label{fig-bazel}}
\end{figure}

@@

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||l|l||l|c|c|c}
\hline
$\!$Build system$\!$& Persistent build information   & Algorithm    & Dependencies & Minimal & Cutoff & Cloud$\!$\\\hline
$\!$\Make       $\!$& File modification times        & Topological  & Static       & Yes     & No     & No   $\!$\\
$\!$\Excel      $\!$& Dirty cells, calculation chain & Reordering   & Dynamic      & No      & No     & No   $\!$\\
$\!$\Shake      $\!$& Previous dependency graph      & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
$\!$\Bazel      $\!$& All dependency graphs, cache   & Topological  & Static       & Yes     & Yes    & Yes  $\!$\\\hline
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Summary of build system differences.\label{tab-summary}}
\vspace{-6mm}
\end{table}
